`{traveltime}`: an R package to calculate walking and motorised travel time from user-specified locations

### 2025-01-29 paper/paper.qmd
paper - further polishing - get rid of ggplot code

`{traveltime}` provides a spatial interface using object classes from the `terra` package `[@terra]`. It accepts points as `matrix`, `data.frame`, or `SpatVector` class objects; and the area of interest as an extent in the form of a `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` object; and returns the result as a `SpatRaster` object. The package uses `gdistance` [@gdistance2017] to calculate the minimum least-cost-distance for each cell from the points of interest. The travel time is calculated as movement over a 'friction surface' based-on the surfaces generated by @weiss2020global and available from the `malariaAtlas` package [@pfeffer2018malariaatlas]. Rather than house and manipulate these surfaces independently, `{traveltime}` relies internally on `malariaAtlas` to download the surfaces for the area of interest.

@@

Other R packages produce superficially similar though fundamentally different functionality. A gaggle of R packages provide interfaces to the [TravelTime.com](https://www.TravelTime.com) API [@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh]. The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons, which are areas reachable within a given time from a given location. The isochron polygons are most comparable to what the `{traveltime}` package calculates, though each calculation is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than gradation across cells. TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repeated iteration for all combinations of time and points, and additional calculations of the minimum value for each cell from all points. Furthermore, TravelTime.com requires access keys, a paid subscription beyond a short free period, and caps queries, which add considerable friction to the use of this resource.

@@

Now that we have the two items of data that require, the next step is to download a friction surface for our area of interest. We can pass in our basemap `sin`, a `SpatVector`, directly as the `extent`. We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`. (Alternatively, we could use `surface = "motor2020"` for motorised travel if that were of interest.) We're also only interested in walking *on land* so we mask out areas outside of `sin`, that are within the extent of the raster:

@@

Below we plot the friction surface raster `friction_singapore`, with the vector boundary of `sin` as a dashed grey line, and `stations` as grey points (Figure \ref{fig:data}). Higher values of friction indicate more time travelling across a given cell.

```{r plot data, fig.cap = "\\label{fig:data}Friction surface raster of Singapore, showing Singapore boundary in grey, and station locations as grey points."}
library(tidyterra)
library(ggplot2)

@@

With all the data collected, the function `calculate_travel_time` takes the friction surface `friction_singapore` and the locations of interest `stations`, and returns a `SpatRaster` of walking time in minutes to each cell from the nearest station as a `SpatRaster`:

@@

We present the resulting calculated travel times in Figure \ref{fig:result}. Note that the results in `travel_time_sin` include infinite (`Inf`) values. In Figure \ref{fig:data}, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland. These raster cells for these islands appear absent in Figure \ref{fig:result}. Because they are not connected to any cells with a station, the calculated travel time is infinite, and so these cells do not appear in Figure \ref{fig:result}.

```{r plot result, fig.cap = "\\label{fig:result}Map of walking travel time in Singapore, in minutes from nearest MRT or LRT station."}
ggplot() +
  geom_spatraster(
    data = travel_time_sin
  ) +
  scale_fill_viridis_c(
    option = "A",
    direction = -1,
    na.value = "transparent"
  ) +
  theme_minimal() +
  labs(fill = "Minutes") +
  geom_spatvector(
    data = sin,
    fill = "transparent",
    col = "grey20"
  )
```

>>>>

`{traveltime}` provides a spatial interface using object classes from the `terra` package [@terra]. It accepts points as `matrix`, `data.frame`, or `SpatVector` class objects; and the area of interest as an extent in the form of a `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` object; and returns the result as a `SpatRaster` object. The package uses `gdistance` [@gdistance2017] to calculate the minimum least-cost-distance for each cell from the points of interest. The travel time is calculated as movement over a 'friction surface' based-on the surfaces generated by @weiss2020global and available from the `malariaAtlas` package [@pfeffer2018malariaatlas]. Rather than house and manipulate these surfaces independently, `{traveltime}` relies internally on `malariaAtlas` to download the surfaces for the area of interest.

@@

Other R packages provide superficially similar though fundamentally different functionality. A gaggle of R packages provide interfaces to the [TravelTime.com](https://www.TravelTime.com) API [@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh]. The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons --- areas reachable within a given time from a given location. The isochron polygons are most comparable to what the `{traveltime}` package calculates, though each calculation is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than gradation across a landscape for multiple points. TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repeated iteration for all combinations of time and points, plus additional calculation of the minimum value for each cell from all points. Furthermore, TravelTime.com requires access keys, a paid subscription beyond a short free period, and caps queries, which add considerable friction to the use of this resource.

@@

Now that we have the two items of data that require, the next step is to download a friction surface for our area of interest. We can pass in our basemap `sin`, a `SpatVector`, directly as the `extent`. We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`. (Alternatively, we could use `surface = "motor2020"` for motorised travel if that were of interest.) We're also only interested in walking *on land* so we then mask out areas outside of `sin`, that are within the extent of the friction surface raster:

@@

Below we plot the friction surface raster `friction_singapore`, with the vector boundary `sin` as a grey line, and `stations` as grey points (Figure \ref{fig:data}). Higher values of friction indicate more time travelling across a given cell.

```{r plot data, echo=FALSE, fig.cap = "\\label{fig:data}Friction surface raster of Singapore, showing Singapore boundary in grey, and station locations as grey points."}
library(tidyterra)
library(ggplot2)

@@

With all the data collected, the function `calculate_travel_time` takes the friction surface `friction_singapore` and the locations of interest `stations`, and returns a `SpatRaster` of walking time in minutes to each cell from the nearest station:

@@

We present the resulting calculated travel times in Figure \ref{fig:result}. Note that the results in `travel_time_sin` include infinite (`Inf`) values. In Figure \ref{fig:data}, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland. The raster cells for these islands appear absent in Figure \ref{fig:result}. Because they are not connected to any cells with a station, the calculated travel time is infinite, and so these cells do not appear in Figure \ref{fig:result}.

```{r plot result, echo=FALSE,  fig.cap = "\\label{fig:result}Map of walking travel time in Singapore, in minutes from nearest MRT or LRT station."}
ggplot() +
  geom_spatraster(
    data = travel_time_sin
  ) +
  scale_fill_viridis_c(
    option = "A",
    direction = -1,
    na.value = "transparent"
  ) +
  theme_minimal() +
  labs(fill = "Minutes") +
  geom_spatvector(
    data = sin,
    fill = "transparent",
    col = "grey20"
  )
```
